---
title: "scRNA-seq Analysis of Human PBMCs using Seurat"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: show
    fig_width: 10
    fig_height: 6
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,      # show code
  message = FALSE,     # suppress package messages
  warning = FALSE,     # suppress warnings
  cache   = TRUE,      # cache heavy chunks (PCA, UMAP, FindAllMarkers)
  dpi     = 150,
  fig.align = "center",
  fig.path  = "figures/",      # saves all figures here automatically
  dev       = "png"            # format: png, pdf, svg
)

# Global seed for reproducibility
set.seed(42)
```

---

## Overview

This report performs an end-to-end single-cell RNA-seq analysis of the [10x Genomics PBMC 3K dataset](https://www.10xgenomics.com/datasets/3-k-pb-mcs-from-a-healthy-donor-1-standard-1-1-0) using **Seurat v5**. The pipeline covers:

1. Data loading
2. Quality control
3. Normalization & feature selection
4. Scaling
5. Dimensionality reduction (PCA + UMAP)
6. Clustering
7. Marker gene identification
8. Cell type annotation

---

## 1. Load Packages

```{r libraries}
library(Seurat)    # core scRNA-seq toolkit
library(tidyverse) # data wrangling + ggplot2
library(patchwork) # combine multiple plots
library(ggplot2)   # plotting
```

---

## 2. Load Data

The dataset is the standard 10x Genomics PBMC 3K output, consisting of
three files produced by Cell Ranger:

- `matrix.mtx` — sparse UMI count matrix
- `barcodes.tsv` — cell barcodes
- `genes.tsv` — gene names

```{r load-data}
# Read the 10x MEX-format files from disk
# Adjust the path if your folder structure differs
pbmc.data <- Read10X(data.dir = "data/filtered_gene_bc_matrices/hg19/")

# Create a Seurat object
# min.cells = 3    : keep genes detected in at least 3 cells
# min.features = 200 : keep cells with at least 200 detected genes
pbmc <- CreateSeuratObject(
  counts       = pbmc.data,
  project      = "PBMC3K",
  min.cells    = 3,
  min.features = 200
)

pbmc
```

The object contains **`r ncol(pbmc)` cells** and **`r nrow(pbmc)` genes** after initial filtering.

---

## 3. Quality Control

Three metrics are used to identify and remove low-quality cells:

| Metric | Low value → | High value → |
|---|---|---|
| `nFeature_RNA` | Empty droplet / dead cell | Potential doublet |
| `nCount_RNA` | Empty droplet | Potential doublet |
| `percent.mt` | — | Dying / stressed cell |

```{r qc-metrics}
# Calculate mitochondrial gene percentage
# Human mt genes are prefixed "MT-"
pbmc[["percent.mt"]] <- PercentageFeatureSet(pbmc, pattern = "^MT-")

# Preview the QC columns
head(pbmc@meta.data, 5)
```

### QC violin plots (before filtering)

```{r qc-violin, fig.height=5, fig.width=12}
VlnPlot(
  pbmc,
  features = c("nFeature_RNA", "nCount_RNA", "percent.mt"),
  ncol     = 3,
  pt.size  = 0.05
)
```

### QC scatter plots

```{r qc-scatter, fig.width=12, fig.height=5}
p1 <- FeatureScatter(pbmc, feature1 = "nCount_RNA", feature2 = "percent.mt")
p2 <- FeatureScatter(pbmc, feature1 = "nCount_RNA", feature2 = "nFeature_RNA")
p1 + p2
```

### Apply QC thresholds

```{r qc-filter}
# Keep cells with:
#   200–2500 detected genes  (removes empty droplets and likely doublets)
#   < 5% mitochondrial reads (removes dying / stressed cells)
pbmc <- subset(
  pbmc,
  subset = nFeature_RNA > 200 &
           nFeature_RNA < 2500 &
           percent.mt   < 5
)

pbmc
```

> **After QC filtering:** `r ncol(pbmc)` cells retained.

---

## 4. Normalization

Log-normalize each cell's counts to account for differences in sequencing
depth across cells.

```{r normalize}
# Divide counts by total per cell, multiply by 10,000, then log1p-transform
pbmc <- NormalizeData(
  pbmc,
  normalization.method = "LogNormalize",
  scale.factor         = 10000
)
```

---

## 5. Feature Selection (Highly Variable Genes)

Identify the top 2,000 genes that vary most across cells — these carry
biological signal and speed up downstream computation.

```{r hvg, fig.width=10, fig.height=5}
pbmc <- FindVariableFeatures(
  pbmc,
  selection.method = "vst",
  nfeatures        = 2000
)

# Label the top 10 most variable genes
top10 <- head(VariableFeatures(pbmc), 10)

plot1 <- VariableFeaturePlot(pbmc)
plot2 <- LabelPoints(plot = plot1, points = top10, repel = TRUE)
plot2
```

Top 10 highly variable genes: **`r paste(top10, collapse = ", ")`**

---

## 6. Scaling

Z-score each gene across all cells so that highly expressed genes do not
dominate PCA. Mitochondrial percentage is regressed out as a technical
confounder.

```{r scale}
all.genes <- rownames(pbmc)

pbmc <- ScaleData(
  pbmc,
  features        = all.genes,
  vars.to.regress = "percent.mt"
)
```

---

## 7. Principal Component Analysis

```{r pca}
pbmc <- RunPCA(
  pbmc,
  features = VariableFeatures(object = pbmc),
  npcs     = 50
)
```

### Top gene loadings per PC

```{r pca-loadings, fig.width=10, fig.height=5}
VizDimLoadings(pbmc, dims = 1:2, reduction = "pca")
```

### PCA scatter

```{r pca-scatter, fig.width=7, fig.height=6}
DimPlot(pbmc, reduction = "pca")
```

### Elbow plot — choosing dimensionality

```{r elbow, fig.width=8, fig.height=4}
# Look for the "elbow" where variance gain levels off
# For PBMC3K this is typically around PC 10
ElbowPlot(pbmc, ndims = 50)
```

> **Decision:** Top **10 PCs** capture the major axes of variation and are
> used for all downstream steps.

---

## 8. Clustering

Graph-based clustering using a KNN graph (FindNeighbors) followed by
the Louvain community-detection algorithm (FindClusters).

```{r cluster}
pbmc <- FindNeighbors(pbmc, dims = 1:10)

pbmc <- FindClusters(
  pbmc,
  resolution = 0.5   # higher = more clusters; 0.5 is standard for PBMCs
)

# Cell counts per cluster
table(Idents(pbmc))
```

---

## 9. UMAP

UMAP embeds cells in 2D for visualization. It uses the same top PCs as
input. **Note:** UMAP coordinates are for visualization only — do not use
them for statistical tests.

```{r umap}
pbmc <- RunUMAP(
  pbmc,
  dims     = 1:10,
  seed.use = 42
)
```

```{r umap-plot, fig.width=8, fig.height=7}
DimPlot(
  pbmc,
  reduction  = "umap",
  label      = TRUE,
  label.size = 5,
  repel      = TRUE
) + ggtitle("PBMC 3K — Seurat clusters")
```

---

## 10. Marker Gene Identification

For each cluster, identify genes significantly enriched versus all other
clusters combined (Wilcoxon rank-sum test).

```{r markers, cache=TRUE}
pbmc.markers <- FindAllMarkers(
  pbmc,
  only.pos        = TRUE,  # upregulated markers only
  min.pct         = 0.25,  # gene detected in ≥25% of cluster cells
  logfc.threshold = 0.25   # minimum log2 fold-change
)
```

### Top 2 markers per cluster

```{r top-markers}
top_markers <- pbmc.markers %>%
  group_by(cluster) %>%
  slice_max(order_by = avg_log2FC, n = 2)

knitr::kable(top_markers, digits = 3)
```

### Marker heatmap (top 10 per cluster)

```{r heatmap, fig.width=14, fig.height=10}
top10_markers <- pbmc.markers %>%
  group_by(cluster) %>%
  slice_max(order_by = avg_log2FC, n = 10)

DoHeatmap(pbmc, features = top10_markers$gene) + NoLegend()
```

### Dot plot

```{r dotplot, fig.width=14, fig.height=6}
DotPlot(
  pbmc,
  features = unique(top10_markers$gene)
) + RotatedAxis()
```

---

## 11. Cell Type Annotation

Manual annotation using canonical PBMC marker genes.

| Cell type | Key markers |
|---|---|
| Naive CD4 T | IL7R, CCR7 |
| Memory CD4 T | IL7R, S100A4 |
| CD8 T | CD8A |
| NK cell | GNLY, NKG7 |
| B cell | MS4A1, CD79A |
| CD14 Monocyte | CD14, LYZ |
| CD16 Monocyte | FCGR3A, MS4A7 |
| Dendritic cell | FCER1A, CST3 |
| Megakaryocyte | PPBP |

### Canonical markers on UMAP

```{r feature-plots, fig.width=16, fig.height=12}
canonical_markers <- c(
  "IL7R", "CCR7",   # CD4 T / Naive T
  "CD8A",           # CD8 T
  "GNLY", "NKG7",   # NK
  "MS4A1",          # B cell
  "CD14", "LYZ",    # CD14 Mono
  "FCGR3A",         # CD16 Mono
  "FCER1A",         # DC
  "PPBP"            # Megakaryocyte
)

FeaturePlot(
  pbmc,
  features   = canonical_markers,
  ncol       = 4,
  min.cutoff = "q10"
)
```

### Assign cell type labels

```{r annotate}
# Map cluster numbers to cell type names
# Verify against YOUR marker output before finalizing
new.cluster.ids <- c(
  "Naive CD4 T",    # 0
  "CD14 Monocyte",  # 1
  "Memory CD4 T",   # 2
  "B Cell",         # 3
  "CD8 T",          # 4
  "CD16 Monocyte",  # 5
  "NK Cell",        # 6
  "Dendritic Cell", # 7
  "Megakaryocyte"   # 8  (may not appear at res=0.5)
)

names(new.cluster.ids) <- levels(pbmc)
pbmc <- RenameIdents(pbmc, new.cluster.ids)
```

### Final annotated UMAP

```{r umap-annotated, fig.width=10, fig.height=7}
DimPlot(
  pbmc,
  reduction  = "umap",
  label      = TRUE,
  label.size = 4,
  repel      = TRUE,
  pt.size    = 0.5
) +
  ggtitle("PBMC 3K — Cell type annotation") +
  theme(legend.position = "right")
```

---

## 12. Save Outputs

```{r save, cache=FALSE}
# Create output directories
dir.create("figures", showWarnings = FALSE)
dir.create("data",    showWarnings = FALSE)

# Save processed Seurat object
saveRDS(pbmc, file = "data/pbmc3k_processed.rds")

# Save all markers as CSV
write.csv(pbmc.markers, file = "data/all_cluster_markers.csv", row.names = FALSE)

message("Outputs saved to data/ directory.")
```

---

## 13. Session Info

```{r session-info}
sessionInfo()
```
